Measuring causal influence with\\ back-to-back regression

### 2019-05-22 paper/paper.tex
backwards done

If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
%
In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{E} \leftarrow \hat{E} + \text{Diag}(H)$\;
    }
    $\hat{E} \leftarrow \hat{E} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
    \Return{$\hat{E}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

Assuming centered noise $N$,
\begin{align*}
    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| (EX + N)FG \right\|^2 \\
                                            &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
\end{align*}
%
As the noise $N$ tends to zero, the solution to the previous problem is $\hat G = (X E F)^\dagger X$, where ``$\dagger$'' denotes the Moore-Penrose pseudo-inverse.

Next, let $k$ be the number of ones in the causal influence matrix $E$.
%
Without loss of generality, assume that these are the first $k$ variables in $X$.
%
Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
%
Let $K \in \mathbb{R}^{d_x \times k}$ be the diagonal matrix obtained by deleting the right-most $d_x - k$ columns of $E$.
%
Then, $E = KK^\top$, and $XEF = (XK)(K^\top F)$, with both $XK$ and $K^\top$ of rank $k$.

Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $C$ is full rank, $C^\dagger=C' (CC')^{-1}$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK')(KF))^\dagger X\\
&=(FK)^\dagger (XK')'((XK')(XK')')^{-1} X\\
&=(FK)^\dagger KX' (K'XX'K)^{-1} X
\end{aligned}
\end{equation}

Let $XX' = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$. Then, $(K'XX'K)=\Sigma_{1}$ and $(XX')K(K'XX'K)^{-1}=\left(\begin{array}{c}I_{k} \\\hline \Sigma_{2} \Sigma_{1}^{-1}\end{array}\right)$, call it S.

Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=X(FEX)^\dagger =S (FK)^\dagger$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $K (FK)^\dagger$.

Under centred noise, and if the covariance of FN has low condition number (a weak assumption given that N is noise), we have $\left \| GFN\right \| ^2 \approx Var(FN) \left \| G\right \| ^2$, and minimisation of $\left \| X-GFEX\right\| ^2  + \left \| GFN\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda S (FE)^\dagger$, $\lambda$ a positive number below one.

Replacing in the left part of (ref), we have $X-GFEX=X- \lambda S (FK)^\dagger FKK'X= X - \lambda EX=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).

For the right part, we have $GFN=\lambda S (FK)^\dagger FN= \lambda S (FK)^\dagger (FK+F(I-K))K'N= \lambda S EN=  \lambda EN$. (Since F is full rank over E, we have $(FK)^\dagger F(I-K)= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.

@@

The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (FK)^\dagger$.

The second step is straightforward, replacing Y, we have
\begin{equation}
\begin{aligned}
\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\
&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
&=\hat G (FE)
\end{aligned}
\end{equation}

Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} S (FK)^{\dagger}(FKK') = \frac{1}{1+nsr} E$.

As we shall see, the diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E. Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
% \subsection{Asymptotic behaviour - non linear case}

\subsection{Binarizing $\hat{E}$}

@@

In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{todo} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

\section{Experiments}

We perform two experiments to evaluate the performance of BB2R: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
%
We use scikit-learn to implement all of our simulations \citep{sklearn}.

\subsection{Synthetic data}

@@

\section{Related work}
Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
%
In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

@@

>>>>

If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
%
In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
        $E \leftarrow E + \text{Diag}(H)$\;
    }
    \Return{$\hat{E} / m$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

Assuming centered noise $N$,
\begin{align*}
    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| X - (XE + N)FG \right\|^2 \\
                                            &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
\end{align*}
%
As the noise $N$ tends to zero, the solution to the previous problem is $\hat G = (X E F)^\dagger X$, where ``$\dagger$'' denotes the Moore-Penrose pseudo-inverse.

Next, let $k$ be the number of ones in the causal influence matrix $E$.
%
Without loss of generality, assume that these are the first $k$ variables in $X$.
%
Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
%
Let $K \in \mathbb{R}^{k \times n_{x}}$ be the diagonal matrix obtained by deleting the bottom $dn_{x} - k$ lines of $E$.
%
Then, $E = K^\top K$, and $XEF = (XK^\top)(KF)$, with both $XK^\top$ and $KF$ of rank $k$.

Recall the rank factorisation property of the pseudo-inverse \citep{todo}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
\end{aligned}
\end{equation}

Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$. Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.

Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.

Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.

Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).

For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$. (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.

@@

The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.

The second step is straightforward, replacing Y, we have
\begin{equation}
\begin{aligned}
\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
&= EF \hat G 
\end{aligned}
\end{equation}

Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.

%The diagonal of $\hat H$ recovered by back to back regression is a scaled approximation of the diagonal of E. Noise tends reduce the large diagonal elements, whereas the bias introduced by regularisation tends to increase the lowest ones.
% \subsection{Asymptotic behaviour - non linear case}

\subsection{Post-processing of B2BR}

@@

In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

\section{Experiments}
\subsection{Synthetic data}

@@

\section{Related works}
\subsection{Back-to-back regression as a special case of canonical component analysis}
Formula (1) shows that back-to-back regression may be related to canonical component analysis (CCA).
%
In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

@@

\subsection{Related regression methods}
FC: discuss similar methods of regression, that can be used for feature extraction
And maybe have something about other causal discovery methods
